---
title: "CFB Analytics — Data & Model Evaluation"
subtitle: "A polyglot college-football pipeline · CFBD → DuckDB → dbt → R + Python models"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    toc-location: left
    embed-resources: true
    fig-width: 8
    fig-height: 4.5
    df-print: kable
    code-fold: true
    code-summary: "Show code"
engine: knitr
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(gt)
library(scales)

# Rendered from the repo root (execute-dir: project). R reads the warehouse exports as flat
# files — Python owns DuckDB; this is the polyglot contract, not a workaround.
gold    <- function(f) read_csv(file.path("data", "gold", f), show_col_types = FALSE)
result  <- function(f) read_csv(file.path("data", "results", f), show_col_types = FALSE)
dash    <- function(f) read_csv(file.path("data", "dashboard", f), show_col_types = FALSE)

theme_set(theme_minimal(base_size = 12))
PAL <- c(r = "#2C7FB8", py = "#D95F0E", model = "#2C7FB8", market = "#D95F0E")
SEASON <- 2025         # headline season for TEAM-level views (latest available)
# cfbfastR's 2025 play-by-play has incomplete PLAYER attribution (~56% of rushes named vs ~96%
# in 2023-24), so PLAYER leaderboards feature the last fully-attributed season instead.
PLAYER_SEASON <- 2024
```

> **Data:** [CollegeFootballData.com](https://collegefootballdata.com) (CFBD API, free tier).
> Raw data is gitignored and collected within the API's bounded/cached terms. This page evaluates
> **(1) the data** and **(2) the models' accuracy**. Every statistical method is implemented in
> **both R and Python** against the same warehouse feed; their agreement is a built-in correctness
> check (see *R ↔ Python parity*).

::: {.callout-tip appearance="simple"}
🏈 **[Open the interactive Team Comparison →](compare.html)** — pick any two FBS teams and compare
their 2025 season head-to-head (SP+, EPA, efficiency, player leaders) with a neutral-field matchup
projection and 2026 win forecast.

📖 **[Stat Guide →](glossary.html)** — a plain-English data dictionary for every metric, with the
2025 FBS range for each. &nbsp;•&nbsp; 🧪 **[The Models →](models.html)** — how the projections
work and how well they perform: an interactive win-probability curve, a real held-out calibration
plot, baseline comparisons, and the R ↔ Python parity gate.
:::

# Part 1 — Evaluating the data

## Data-quality panel

```{r dq-panel}
counts <- dash("dq_layer_counts.csv")
tests  <- dash("dq_tests.csv")
fresh  <- dash("dq_freshness.csv")

n_pass <- tests |> filter(status == "pass") |> pull(n) |> sum()
n_fail <- tests |> filter(status %in% c("fail", "error")) |> pull(n) |> sum()

layer_summary <- counts |>
  mutate(layer = factor(layer, levels = c("staging", "silver", "gold", "snapshots"))) |>
  group_by(layer) |>
  summarise(tables = n(), rows = sum(row_count), .groups = "drop")

layer_summary |>
  gt() |>
  tab_header(
    title = "Medallion warehouse",
    subtitle = sprintf("dbt data tests: %d passed, %d failed · sources fresh as of %s",
                       n_pass, n_fail, format(max(fresh$last_loaded), "%Y-%m-%d"))
  ) |>
  fmt_integer(columns = c(tables, rows)) |>
  cols_label(layer = "Layer", tables = "Tables", rows = "Rows") |>
  data_color(columns = layer, palette = "Blues", alpha = 0.4) |>
  tab_options(table.font.size = px(13))
```

```{r dq-freshness}
fresh |>
  arrange(desc(rows)) |>
  gt() |>
  tab_header(title = "Bronze sources — load lineage") |>
  fmt_integer(columns = rows) |>
  cols_label(source = "Source", last_loaded = "Last loaded", rows = "Rows") |>
  tab_options(table.font.size = px(12))
```

## Team efficiency leaderboard

Opponent-context offensive/defensive **EPA per play** (lower defensive EPA = better), season average.

```{r team-eff}
te <- dash("team_efficiency.csv") |> filter(season == SEASON)

te |>
  arrange(desc(net_epa)) |>
  slice_head(n = 12) |>
  transmute(Team = team, `W%` = win_pct, `Net EPA` = net_epa,
            `Off EPA` = off_epa, `Def EPA` = def_epa) |>
  gt() |>
  tab_header(title = sprintf("Top teams by net EPA/play — %d", SEASON)) |>
  fmt_percent(columns = `W%`, decimals = 0) |>
  fmt_number(columns = c(`Net EPA`, `Off EPA`, `Def EPA`), decimals = 3) |>
  data_color(columns = `Net EPA`, palette = "Greens") |>
  tab_source_note("Net EPA = offensive EPA/play − defensive EPA/play allowed.") |>
  tab_options(table.font.size = px(12))
```

## "Over expected" leaderboards (M2–M4)

The book's residual framework: model the *expected* outcome from game situation; the residual is
skill. **RYOE** (rushing yards over expected, M3) and **CPOE** (completion % over expected, M4).
*(Featuring `r PLAYER_SEASON` — the latest season with complete player attribution in the source;
2025 play-by-play is fully loaded at the team level but names only ~56% of rushes so far.)*

```{r ryoe-cpoe}
plays <- gold("plays_model.csv") |>
  select(play_key, season, rusher_player_name, passer_player_name, offense_team)

ryoe_lb <- result("pred__ryoe__r.csv") |>
  inner_join(plays, by = "play_key") |>
  filter(season == PLAYER_SEASON, !is.na(rusher_player_name)) |>
  group_by(Player = rusher_player_name, Team = offense_team) |>
  summarise(Carries = n(), `RYOE/att` = mean(ryoe), .groups = "drop") |>
  filter(Carries >= 100) |>
  arrange(desc(`RYOE/att`)) |>
  slice_head(n = 10)

cpoe_lb <- result("pred__cpoe__r.csv") |>
  inner_join(plays, by = "play_key") |>
  filter(season == PLAYER_SEASON, !is.na(passer_player_name)) |>
  group_by(Player = passer_player_name, Team = offense_team) |>
  summarise(Attempts = n(), `CPOE` = mean(cpoe), .groups = "drop") |>
  filter(Attempts >= 200) |>
  arrange(desc(`CPOE`)) |>
  slice_head(n = 10)

gt(ryoe_lb) |>
  tab_header(title = "Rushing Yards Over Expected", subtitle = sprintf("%d · ≥100 carries", PLAYER_SEASON)) |>
  fmt_number(columns = `RYOE/att`, decimals = 2) |>
  data_color(columns = `RYOE/att`, palette = "Blues") |>
  tab_options(table.font.size = px(12))
```

```{r cpoe-lb}
gt(cpoe_lb) |>
  tab_header(title = "Completion % Over Expected", subtitle = sprintf("%d · ≥200 attempts", PLAYER_SEASON)) |>
  fmt_percent(columns = CPOE, decimals = 1) |>
  data_color(columns = CPOE, palette = "Blues") |>
  tab_options(table.font.size = px(12))
```

## Metric stability (M1) — is it skill or noise?

Split-half reliability (odd vs even plays) of rushing yards-per-carry. A tight diagonal = a
reliable, skill-driven metric; a loose cloud = mostly noise (and a candidate for shrinkage, M7).

```{r stability-plot}
pr <- gold("player_rushing.csv") |>
  filter(season == PLAYER_SEASON, attempts_odd >= 40, attempts_even >= 40)

r_split <- cor(pr$ypc_odd_plays, pr$ypc_even_plays, use = "complete.obs")

ggplot(pr, aes(ypc_odd_plays, ypc_even_plays)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(alpha = 0.5, colour = PAL["model"], size = 1.8) +
  geom_smooth(method = "lm", se = FALSE, colour = PAL["market"], linewidth = 0.8) +
  labs(
    title = sprintf("Rushing YPC split-half reliability, %d  (r = %.2f)", PLAYER_SEASON, r_split),
    subtitle = "Each point = a rusher (≥40 carries in each half). Loose spread ⇒ YPC is noisy.",
    x = "Yards/carry — odd plays", y = "Yards/carry — even plays") +
  coord_equal()
```

```{r stability-table}
bind_rows(result("metrics__stability__r.csv"), result("metrics__stability__py.csv")) |>
  filter(metric %in% c("split_half_reliability_ypc", "year_over_year_r_ypc")) |>
  mutate(metric = recode(metric,
                         split_half_reliability_ypc = "Split-half reliability (Spearman-Brown)",
                         year_over_year_r_ypc = "Year-over-year r")) |>
  select(metric, basis, n, value, language) |>
  pivot_wider(names_from = language, values_from = value) |>
  gt() |>
  tab_header(title = "Reliability of rushing YPC", subtitle = "R and Python agree") |>
  fmt_number(columns = c(r, py), decimals = 3) |>
  cols_label(metric = "Metric", basis = "Season(s)", n = "n",
             r = "R", py = "Python") |>
  tab_options(table.font.size = px(12))
```

## Team archetypes (M6) — PCA + k-means

Each team-season's 7-stat style profile reduced to 2 principal components, then clustered into
five archetypes. Efficient, complete teams sit to one side; struggling or one-dimensional teams to
the other.

```{r archetypes}
arch <- result("pred__archetype__r.csv") |> mutate(cluster = factor(cluster))
prof <- gold("team_profile.csv")
arch <- arch |> left_join(prof, by = c("team", "season"))

# label a few notable 2024 teams so the map is legible
notable <- c("Ohio State", "Oregon", "Georgia", "Florida State", "Miami", "Army")
labels24 <- arch |> filter(season == SEASON, team %in% notable)

ggplot(arch, aes(pc1, pc2, colour = cluster)) +
  geom_point(alpha = 0.55, size = 1.8) +
  geom_text(data = labels24, aes(label = team), colour = "black", size = 3,
            vjust = -0.8, show.legend = FALSE) +
  scale_colour_brewer(palette = "Set1", name = "Archetype") +
  labs(title = "Team archetypes — principal components of team style (2023–25)",
       subtitle = "PC1 ≈ overall efficiency; nearby teams play alike. Points = team-seasons.",
       x = "PC1", y = "PC2")
```

```{r archetype-profiles}
arch |>
  group_by(Archetype = cluster) |>
  summarise(n = n(), `Off EPA` = mean(off_epa_play), `Def EPA` = mean(def_epa_play),
            `Run rate` = mean(rush_rate), Pace = mean(off_pace), .groups = "drop") |>
  gt() |>
  tab_header(title = "Archetype centroids",
             subtitle = "Mean style profile per cluster (Def EPA lower = better)") |>
  fmt_number(columns = c(`Off EPA`, `Def EPA`), decimals = 3) |>
  fmt_number(columns = c(`Run rate`, Pace), decimals = 2) |>
  tab_options(table.font.size = px(12))
```

## Expected points by field position — the EPA foundation

Every efficiency metric here is built on **EPA** (expected points added). This is the expected-
points surface it rests on: the mean expected points for a first-and-10 (and later downs) at each
spot on the field — from ~6.4 points at the goal line down to *negative* when backed up near your
own end zone. The monotonic, football-sensible shape is the sanity check for the whole EPA layer.

```{r ep-surface}
eps <- gold("ep_surface.csv") |> mutate(down = factor(down))

ggplot(eps, aes(yards_to_goal, expected_points, colour = down)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_line(linewidth = 0.9) +
  scale_x_reverse() +
  scale_colour_brewer(palette = "Set1", name = "Down") +
  labs(title = "Expected points by field position and down (2023–25)",
       subtitle = "x reversed: goal line at right. EP falls with distance and later downs.",
       x = "Yards to opponent's goal", y = "Expected points")
```

# Part 2 — Evaluating the models

## The game win-probability model

Logistic model of P(home win) from leakage-safe entering-form efficiency differentials. Trained on
the earlier seasons (2023–24) and evaluated on the **sealed 2025 holdout** — a season it never saw —
against a home-field-naive baseline and the **betting market** (the bar to clear).

```{r game-metrics}
gm <- bind_rows(result("metrics__game__r.csv"), result("metrics__game__py.csv")) |>
  filter(language == "r")
val <- function(m) gm |> filter(metric == m) |> pull(value) |> first()

tibble(
  Metric = c("Brier score", "Log loss", "AUC", "Accuracy"),
  Model  = c(val("brier"), val("log_loss"), val("auc"), val("accuracy")),
  `Market line` = c(val("brier_market_line"), NA, NA, val("accuracy_market_line")),
  `Naive (home-field)` = c(val("brier_naive_homefield"), NA, NA, NA)
) |>
  gt() |>
  tab_header(title = "2025 holdout accuracy",
             subtitle = "Lower Brier/log-loss = better; the model approaches the market using on-field efficiency only") |>
  fmt_number(columns = c(Model, `Market line`, `Naive (home-field)`), decimals = 3) |>
  sub_missing(missing_text = "—") |>
  tab_options(table.font.size = px(12))
```

```{r calibration}
gp <- result("gamepred__game__r.csv")

calib <- gp |>
  mutate(bin = cut(home_win_prob, breaks = seq(0, 1, 0.1), include.lowest = TRUE)) |>
  group_by(bin) |>
  summarise(pred = mean(home_win_prob), obs = mean(home_won), n = n(), .groups = "drop")

ggplot(calib, aes(pred, obs)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_line(colour = PAL["model"], linewidth = 0.8) +
  geom_point(aes(size = n), colour = PAL["model"]) +
  scale_size_area(max_size = 8, guide = "none") +
  scale_x_continuous(labels = percent, limits = c(0, 1)) +
  scale_y_continuous(labels = percent, limits = c(0, 1)) +
  labs(title = "Calibration — predicted vs observed home-win rate (2025 holdout)",
       subtitle = "On the diagonal = well-calibrated probabilities. Point size = games in bin.",
       x = "Predicted P(home win)", y = "Observed home-win rate") +
  coord_equal()
```

```{r brier-by-week}
bw <- gp |>
  group_by(week) |>
  summarise(Model = mean((home_won - home_win_prob)^2),
            Market = mean((home_won - market_win_prob)^2),
            n = n(), .groups = "drop") |>
  pivot_longer(c(Model, Market), names_to = "source", values_to = "brier")

ggplot(bw, aes(week, brier, colour = source)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  scale_colour_manual(values = c(Model = unname(PAL["model"]), Market = unname(PAL["market"])),
                      name = NULL) +
  labs(title = "Brier score by week — model vs betting market (2025)",
       subtitle = "The market is hard to beat; the efficiency model tracks it closely.",
       x = "Week", y = "Brier score")
```

## Forecasting a new season before it starts

The in-season model needs a few weeks of *this year's* results. To predict **openers** — or a whole
schedule the day it drops — a **preseason priors model** uses only *last season's* carryover (SP+
rating, net EPA, win rate). It's genuinely useful, but the table below quantifies the honest cost of
having no current-year form: preseason forecasts are meaningfully less certain. This is the model
that would score a 2026 schedule the moment it's loaded.

```{r priors-vs-inseason}
prio <- result("metrics__priors__r.csv")
game <- result("metrics__game__r.csv") |> filter(language == "r")
pick <- function(d, m) d |> filter(metric == m) |> pull(value) |> first()

tibble(
  Metric = c("Brier (lower better)", "AUC (higher better)", "Accuracy"),
  `In-season model` = c(pick(game, "brier"), pick(game, "auc"), pick(game, "accuracy")),
  `Preseason priors` = c(pick(prio, "brier"), pick(prio, "auc"), pick(prio, "accuracy")),
  `Naive baseline` = c(pick(prio, "brier_naive"), NA, NA)
) |>
  gt() |>
  tab_header(title = "In-season vs preseason forecasting (2025 holdout)",
             subtitle = "Preseason (prior-year priors only) beats a naive guess but trails the in-season model") |>
  fmt_number(columns = c(`In-season model`, `Preseason priors`, `Naive baseline`), decimals = 3) |>
  sub_missing(missing_text = "—") |>
  tab_options(table.font.size = px(12))
```

## Model coefficients & fit (M3–M5)

```{r model-metrics}
mt <- bind_rows(
  result("metrics__ryoe__r.csv"), result("metrics__cpoe__r.csv"),
  result("metrics__poisson__r.csv"))

mt |>
  filter(metric %in% c("rmse", "r_squared", "brier", "brier_baseline",
                        "log_loss", "dispersion", "aic_poisson")) |>
  mutate(value = round(value, 4)) |>
  select(Model = model, Metric = metric, Value = value) |>
  gt() |>
  tab_header(title = "Headline fit metrics by model") |>
  tab_options(table.font.size = px(12)) |>
  data_color(columns = Model, palette = "Pastel1", alpha = 0.5)
```

## Multilevel shrinkage (M7) — regression to the mean

M1 showed rushing efficiency is noisy; a random-intercept model (`ryoe ~ 1 + (1 | rusher)`) acts
on it. Partial pooling pulls each rusher's raw RYOE toward the league mean — hardest for
small-sample rushers. Below, the raw averages (grey) fan out wildly at low volume while the
shrunken estimates (blue) stay disciplined; only high-volume backs keep a large signal.

```{r shrinkage}
sh <- result("pred__shrinkage__r.csv")

sh_long <- sh |>
  select(carries, Raw = raw_ryoe, Shrunk = shrunk_ryoe) |>
  tidyr::pivot_longer(c(Raw, Shrunk), names_to = "estimate", values_to = "ryoe")

ggplot(sh_long, aes(carries, ryoe, colour = estimate)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_point(alpha = 0.4, size = 1.3) +
  scale_x_log10() +
  scale_colour_manual(values = c(Raw = "grey60", Shrunk = unname(PAL["model"])), name = NULL) +
  labs(title = "Shrinkage of rushing RYOE — raw vs partially-pooled estimate",
       subtitle = "Grey = raw per-rusher average; blue = shrunken (trusted) estimate. x = carries (log).",
       x = "Carries (log scale)", y = "RYOE per carry")
```

```{r shrinkage-icc}
icc <- result("metrics__shrinkage__r.csv") |> filter(metric == "icc") |> pull(value)
```

Only **`r sprintf("%.1f%%", 100 * icc)`** of RYOE variance is *between* rushers (the intra-class
correlation) — the rest is play-to-play noise. That tiny ICC is exactly why the pooling is so
aggressive, and it quantitatively confirms the M1 stability finding.

## Do teams beat their recruiting? (M8)

Recruiting rankings here were collected by **ethically scraping Wikipedia** (CC BY-SA, bot-friendly)
— only the *team-level* ranking table, never individual (often minor) recruits — and were validated
to match CFBD's sanctioned recruiting API exactly. Regressing each team's on-field production (SP+
rating) on its recruiting-class rank shows the expected relationship (better classes → better teams),
but the **residuals** are the story: who out- or under-performed the talent they signed.

```{r recruiting}
rec <- result("pred__recruiting__r.csv")
extremes <- bind_rows(slice_max(rec, performance_vs_recruiting, n = 3),
                      slice_min(rec, performance_vs_recruiting, n = 3))

ggplot(rec, aes(recruiting_rank_247, sp_rating)) +
  geom_smooth(method = "lm", se = FALSE, colour = "grey55", linewidth = 0.8) +
  geom_point(aes(colour = performance_vs_recruiting), size = 2.6) +
  geom_text(data = extremes, aes(label = paste0(team, " '", substr(season, 3, 4))),
            size = 3, vjust = -0.9, colour = "black") +
  scale_colour_gradient2(low = "#B2182B", mid = "grey80", high = "#2166AC",
                         name = "Over /\nunder") +
  labs(title = "On-field production vs recruiting-class rank (top-25 classes, 2023–25)",
       subtitle = "Above the line = beating their recruiting; below = underachieving.",
       x = "Recruiting-class rank (1 = best)", y = "SP+ rating (higher = better)")
```

```{r recruiting-table}
rec |>
  slice_max(performance_vs_recruiting, n = 6) |>
  transmute(Team = team, Season = season, `Recruit rank` = recruiting_rank_247,
            `SP+` = sp_rating, `Beat recruiting by` = performance_vs_recruiting) |>
  gt() |>
  tab_header(title = "Biggest overachievers vs recruiting",
             subtitle = "Production well above what their signed talent predicted") |>
  fmt_number(columns = c(`SP+`, `Beat recruiting by`), decimals = 1) |>
  data_color(columns = `Beat recruiting by`, palette = "Blues") |>
  tab_options(table.font.size = px(12))
```

# Part 3 — 2026 season outlook (a live forecast)

Everything above measures the past; this looks **forward**. Taking the real **2026 schedule** (pulled
from CFBD, none of it played yet) and each team's 2025 carryover, the preseason priors model scores
every game, and the per-game probabilities add up to a **projected win total** for each team. Read
this as an informed *prior*, not a prophecy: it knows how teams finished 2025 and who they play in
2026, but not transfers, coaching changes, or breakout freshmen — and preseason forecasts are the
least certain kind (the accuracy table above quantifies that).

```{r forecast-load}
have_forecast <- file.exists(file.path("data", "gold", "forecast_2026.csv")) &&
  file.exists(file.path("data", "gold", "forecast_2026_teams.csv"))
```

## Projected 2026 win totals

Expected wins summed over each team's **FBS-vs-FBS** games (cupcake FCS games excluded, so most
teams' real totals run ~1 higher). It blends *team quality* and *schedule*: a strong team with a soft
slate projects high; a good team in a brutal division less so.

```{r projected-wins}
#| eval: !expr have_forecast
teams <- readr::read_csv(file.path("data", "gold", "forecast_2026_teams.csv"),
                         show_col_types = FALSE)

top20 <- teams |> slice_head(n = 20)
ggplot(top20, aes(x = projected_wins, y = reorder(team, projected_wins))) +
  geom_col(fill = PAL["model"], alpha = 0.85) +
  geom_text(aes(label = sprintf("%.1f", projected_wins)), hjust = -0.15, size = 3) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.08))) +
  labs(title = "Projected 2026 win totals — top 20 (FBS-vs-FBS games only)",
       subtitle = "Expected wins = sum of the model's per-game win probabilities.",
       x = "Projected wins", y = NULL)
```

```{r projected-wins-table}
#| eval: !expr have_forecast
teams |>
  slice_head(n = 25) |>
  transmute(`#` = projected_rank, Team = team, `FBS games` = games,
            `Proj. wins` = projected_wins, `Proj. losses` = projected_losses,
            `Avg win prob` = avg_win_prob) |>
  gt() |>
  tab_header(title = "2026 projected win totals — top 25",
             subtitle = "Preseason priors model applied to the real 2026 schedule") |>
  fmt_number(columns = c(`Proj. wins`, `Proj. losses`), decimals = 1) |>
  fmt_percent(columns = `Avg win prob`, decimals = 0) |>
  data_color(columns = `Proj. wins`, palette = "Greens") |>
  tab_options(table.font.size = px(12))
```

## Opening weekend (week 1)

The model's closest calls — genuine toss-ups where it has little conviction — and the games it's most
sure about.

```{r week1-tossups}
#| eval: !expr have_forecast
fc1 <- readr::read_csv(file.path("data", "gold", "forecast_2026.csv"),
                       show_col_types = FALSE) |> filter(week == 1)

fc1 |>
  mutate(edge = abs(home_win_prob - 0.5)) |>
  slice_min(edge, n = 10) |>
  transmute(Matchup = paste(away_team, "@", home_team),
            `Home win prob` = home_win_prob, `Model leans` = favored_team) |>
  gt() |>
  tab_header(title = "Week 1 — the model's true toss-ups",
             subtitle = "≈50% = a coin flip even to the model") |>
  fmt_percent(columns = `Home win prob`, decimals = 0) |>
  tab_options(table.font.size = px(12))
```

```{r week1-locks}
#| eval: !expr have_forecast
fc1 |>
  slice_max(favored_win_prob, n = 8) |>
  transmute(Matchup = paste(away_team, "@", home_team),
            Favorite = favored_team, `Win prob` = favored_win_prob) |>
  gt() |>
  tab_header(title = "Week 1 — the model's most confident picks") |>
  fmt_percent(columns = `Win prob`, decimals = 0) |>
  data_color(columns = `Win prob`, palette = "Blues") |>
  tab_options(table.font.size = px(12))
```

```{r no-forecast-note}
#| eval: !expr (!have_forecast)
#| output: asis
cat("> _2026 forecast not generated. Run `uv run python -m cfb_analytics.forecast 2026`",
    "then re-render._")
```

## R ↔ Python parity — the built-in correctness check

Every method is fit independently in R and Python on the identical feed. Because the fits are
mathematically the same (OLS, IRLS GLM, Poisson, logistic MLE), the coefficients **must** agree.
The pipeline's `load_results` step fails the build if any term diverges beyond tolerance.

```{r parity}
coef_r  <- bind_rows(lapply(
  list.files("data/results", pattern = "^coef__.*__r\\.csv$", full.names = TRUE),
  read_csv, show_col_types = FALSE)) |> select(model, term, estimate_r = estimate)
coef_py <- bind_rows(lapply(
  list.files("data/results", pattern = "^coef__.*__py\\.csv$", full.names = TRUE),
  read_csv, show_col_types = FALSE)) |> select(model, term, estimate_py = estimate)

parity <- inner_join(coef_r, coef_py, by = c("model", "term")) |>
  mutate(`|Δ|` = abs(estimate_r - estimate_py))

max_diff <- max(parity$`|Δ|`)

parity |>
  arrange(desc(`|Δ|`)) |>
  slice_head(n = 12) |>
  gt() |>
  tab_header(title = "R vs Python coefficient agreement",
             subtitle = sprintf("%d coefficients across %d models · max |Δ| = %.2e",
                                nrow(parity), n_distinct(parity$model), max_diff)) |>
  fmt_number(columns = c(estimate_r, estimate_py), decimals = 4) |>
  fmt_scientific(columns = `|Δ|`, decimals = 1) |>
  cols_label(model = "Model", term = "Term", estimate_r = "R", estimate_py = "Python") |>
  data_color(columns = `|Δ|`, palette = c("#1A9850", "#1A9850")) |>
  tab_options(table.font.size = px(12))
```

The unsupervised and mixed-effects methods (M6, M7) don't reduce to a single comparable
coefficient — their fits are sign- or label-ambiguous — so agreement is measured label-invariantly:

```{r parity-m67}
agree <- bind_rows(
  result("metrics__archetype_parity__py.csv"),
  result("metrics__shrinkage_parity__py.csv")) |>
  filter(metric %in% c("pc1_abs_corr_r_vs_py", "cluster_adjusted_rand_index",
                       "blup_corr_r_vs_py")) |>
  mutate(metric = recode(metric,
    pc1_abs_corr_r_vs_py = "M6 · PC1 |correlation|",
    cluster_adjusted_rand_index = "M6 · cluster adjusted Rand index",
    blup_corr_r_vs_py = "M7 · shrunken-estimate correlation")) |>
  select(Check = metric, Agreement = value)

gt(agree) |>
  tab_header(title = "R vs Python — M6/M7 agreement",
             subtitle = "1.0 = identical (PCA up to sign; clusters up to labelling)") |>
  fmt_number(columns = Agreement, decimals = 3) |>
  tab_options(table.font.size = px(12))
```

---

*Built with R (`gt`, `ggplot2`), Python (`duckdb`, `pandas`), dbt, and DuckDB. Models from
Eager & Erickson, *Football Analytics with Python & R*, applied to CFB. Data:
CollegeFootballData.com.*
